---
title: "Project for Reproducible Research"
author: "Nadim Muhammad, Michał Kulbat"
date: "26/05/2023"
format: html
editor: visual
output: 
  html_document:
    theme: spacelab
    highlight: tango
    toc: true
    toc_float:
      collapsed: false
      smooth_scroll: true
---

# 1. Importing libraries and loading dataset

(install Rtools43 first: https://cran.r-project.org/bin/windows/Rtools/rtools43/rtools.html)

```{r warning=FALSE}
library(devtools)
```

```{r warning=FALSE}
#| eval: false
devtools::install_github("dongyuanwu/RSBID")
```

```{r warning=FALSE}
#| warning: false
#<<<<<<< HEAD
library(xts)
library(lmtest)
library(quantmod)
library(dplyr)
library(tidyr)
library(fUnitRoots)
library(vars)
library(tseries)
library(aTSA)
library(car)
library(seasonal)
library(ggplot2) #Tables
library(forecast)
library(kableExtra)
library(formattable)
library(Metrics)
library(TSEwgt)
#library(RSBID) #We will need SMOTE_NC from it
library(tidyverse)
#=======
library(dplyr) #data manipulation
library(formattable) #data manipulation
library(Metrics) #metrics
#library(MLmetrics) #metrics for machine learning (similar to Metrics library)
#library(e1071) #svm
#library(xgboost) #xgboost model
#library(caret) #dummy variables, precision metrics, knn
#library(RSBID) #We will need SMOTE_NC from it
library(tidyverse) #collection of packages for data science
#>>>>>>> 1b3e2b3bbf38c00b31d322a3bd672e29a047b39a
```

Data loading:

```{r}
rain_df <- read.csv("weatherAUS.csv")
```

```{r}
tail(rain_df)
```

# 2. Exploration and preprocessing

## 2.1. RainTomorrow distribution

```{r}
# Check the original shape of the dataset (rows, columns)

dim(rain_df)
```


```{r}
# Check for missing data in target variable

sum(is.na(rain_df$RainTomorrow))
```

#```{r}
# Drop the data not including target variable since using them may later negativley influence the model

rain_df <- rain_df %>% drop_na(c("RainTomorrow"))
dim(rain_df)
```

```{r}
# Check distribution, look if balanced/imbalanced

table(rain_df$RainTomorrow)
```

```{r}
# Visualise distibution to check the impalance

barplot(table(rain_df$RainTomorrow)) 
```

There is a class imbalance in **RainTommorow** - Our target variable is imbalanced.

The data will be transformed from categorical into a numerical for process convinience where 1 will mean that it will rain tomorrow and 0 that it will not rain

```{r echo=TRUE, results='hide'}
# Replace objects Yes with 1 and No with 0

rain_df["RainTomorrow"][rain_df["RainTomorrow"] == "Yes"] <- 1
rain_df["RainTomorrow"][rain_df["RainTomorrow"] == "No"] <- 0
rain_df$RainTomorrow <- as.numeric(rain_df$RainTomorrow)
rain_df["RainTomorrow"]
```

## 2.2. Data tpyes

```{r}
# Check for data type and data amount in columns.

str(rain_df)
```

We can see that there are 23 variables in which:

**1 Date:** Date

**7 Integer:** Cloud9am, Cloud3pm, WindSpeed9am, WindSpeed3pm, Humidity9am, Humidity3pm, WindGustSpeed.

**5 Character:** Location, WindGustDir, WindDir9am, WindDir3pm ,RainToday.

**9 Numerical:** Temp9am, Temp3pm, Pressure9am, Pressure3pm, MinTemp, MaxTemp, Rainfall, Evaporation, Sunshine, RainTomorrow.  

```{r}
# Check for unique data for Categorical in order to decide weather one hot encoding is viable option.

n_distinct(rain_df$Date)
n_distinct(rain_df$Location)
n_distinct(rain_df$MinTemp)
n_distinct(rain_df$MaxTemp)
n_distinct(rain_df$Rainfall)
n_distinct(rain_df$Evaporation)
n_distinct(rain_df$Sunshine)
n_distinct(rain_df$WindGustDir)
n_distinct(rain_df$WindGustDir)
n_distinct(rain_df$WindGustSpeed)
n_distinct(rain_df$WindDir9am)
n_distinct(rain_df$WindDir3pm)
n_distinct(rain_df$WindSpeed9am)
n_distinct(rain_df$WindSpeed3pm)
n_distinct(rain_df$Humidity9am)
n_distinct(rain_df$Humidity3pm)
n_distinct(rain_df$Pressure9am)
n_distinct(rain_df$Pressure3pm)
n_distinct(rain_df$Cloud9am)
n_distinct(rain_df$Cloud3pm)
n_distinct(rain_df$Temp9am)
n_distinct(rain_df$Temp3pm)
n_distinct(rain_df$RainToday)
n_distinct(rain_df$RainTomorrow)
```

**Character** data have to be further transofrmed in order to be used in the model.

## 2.3. Search for invalid and missing data

```{r}
# Look for invalid data

summary(rain_df)
```

```{r}
# Look for missing data

sum(is.na(rain_df$Date))
sum(is.na(rain_df$Location))
sum(is.na(rain_df$MinTemp))
sum(is.na(rain_df$MaxTemp))
sum(is.na(rain_df$Rainfall))
sum(is.na(rain_df$Evaporation))
sum(is.na(rain_df$Sunshine))
sum(is.na(rain_df$WindGustDir))
sum(is.na(rain_df$WindGustDir))
sum(is.na(rain_df$WindGustSpeed))
sum(is.na(rain_df$WindDir9am))
sum(is.na(rain_df$WindDir3pm))
sum(is.na(rain_df$WindSpeed9am))
sum(is.na(rain_df$WindSpeed3pm))
sum(is.na(rain_df$Humidity9am))
sum(is.na(rain_df$Humidity3pm))
sum(is.na(rain_df$Pressure9am))
sum(is.na(rain_df$Pressure3pm))
sum(is.na(rain_df$Cloud9am))
sum(is.na(rain_df$Cloud3pm))
sum(is.na(rain_df$Temp9am))
sum(is.na(rain_df$Temp3pm))
sum(is.na(rain_df$RainToday))
sum(is.na(rain_df$RainTomorrow))
```

There seems to be a lot of missing data epsepcially in: **Evaporation, Sunshine, Cloud9am, and Cloud3pm**

It seems that variables **Rainfall, Evaporation, WindGustSpeed, WindSpeed9am, WindSpeed3pm** have very high difference betwen the mean and max value we will investigate this data using box plot.

## 2.4. Handling missing data

IMPORTANT: We decided to limit our test and training data to those where 'Sunshine' variable is not null in order to reduce the amount of data used and to lower the number of null values in whole dataset

```{r}
rain_df <- rain_df %>% drop_na(c("Sunshine"))
```

```{r}
# Check for missing data

sum(is.na(rain_df$Date)) 
sum(is.na(rain_df$Location))
sum(is.na(rain_df$MinTemp))
sum(is.na(rain_df$MaxTemp))
sum(is.na(rain_df$Rainfall))
sum(is.na(rain_df$Evaporation))
sum(is.na(rain_df$Sunshine))
sum(is.na(rain_df$WindGustDir))
sum(is.na(rain_df$WindGustDir))
sum(is.na(rain_df$WindGustSpeed))
sum(is.na(rain_df$WindDir9am))
sum(is.na(rain_df$WindDir3pm))
sum(is.na(rain_df$WindSpeed9am))
sum(is.na(rain_df$WindSpeed3pm))
sum(is.na(rain_df$Humidity9am))
sum(is.na(rain_df$Humidity3pm))
sum(is.na(rain_df$Pressure9am))
sum(is.na(rain_df$Pressure3pm))
sum(is.na(rain_df$Cloud9am))
sum(is.na(rain_df$Cloud3pm))
sum(is.na(rain_df$Temp9am))
sum(is.na(rain_df$Temp3pm))
sum(is.na(rain_df$RainToday))
sum(is.na(rain_df$RainTomorrow))
```

```{r}
# fill missing data for numeric attributes with mean and median (if there are outliers [max values are much bigger than mean values] in our attribute we will use median)

rain_df$Temp3pm[is.na(rain_df$Temp3pm)] <- mean(rain_df$Temp3pm, na.rm = TRUE)
rain_df$Temp9am[is.na(rain_df$Temp9am)] <- mean(rain_df$Temp9am, na.rm = TRUE)
rain_df$Cloud3pm[is.na(rain_df$Cloud3pm)] <- mean(rain_df$Cloud3pm, na.rm = TRUE)
rain_df$Cloud9am[is.na(rain_df$Cloud9am)] <- mean(rain_df$Cloud9am, na.rm = TRUE)
rain_df$Pressure3pm[is.na(rain_df$Pressure3pm)] <- mean(rain_df$Pressure3pm, na.rm = TRUE)
rain_df$Pressure9am[is.na(rain_df$Pressure9am)] <- mean(rain_df$Pressure9am, na.rm = TRUE)
rain_df$Humidity3pm[is.na(rain_df$Humidity3pm)] <- mean(rain_df$Humidity3pm, na.rm = TRUE)
rain_df$Humidity9am[is.na(rain_df$Humidity9am)] <- mean(rain_df$Humidity9am, na.rm = TRUE)
rain_df$WindSpeed3pm[is.na(rain_df$WindSpeed3pm)] <- median(rain_df$WindSpeed3pm, na.rm = TRUE)
rain_df$WindSpeed9am[is.na(rain_df$WindSpeed9am)] <- median(rain_df$WindSpeed9am, na.rm = TRUE)
rain_df$WindGustSpeed[is.na(rain_df$WindGustSpeed)] <- median(rain_df$WindGustSpeed, na.rm = TRUE)
rain_df$Sunshine[is.na(rain_df$Sunshine)] <- mean(rain_df$Sunshine, na.rm = TRUE)
rain_df$Evaporation <- ifelse(is.na(rain_df$Evaporation), median(rain_df$Evaporation, na.rm = TRUE), rain_df$Evaporation)
rain_df$Rainfall <- ifelse(is.na(rain_df$Rainfall), median(rain_df$Rainfall, na.rm = TRUE), rain_df$Rainfall)
rain_df$MaxTemp <- ifelse(is.na(rain_df$MaxTemp), mean(rain_df$MaxTemp, na.rm = TRUE), rain_df$MaxTemp)
rain_df$MinTemp <- ifelse(is.na(rain_df$MinTemp), mean(rain_df$MinTemp, na.rm = TRUE), rain_df$MinTemp)
```

```{r}
# Look for missing data again

sum(is.na(rain_df$Date)) 
sum(is.na(rain_df$Location))
sum(is.na(rain_df$MinTemp))
sum(is.na(rain_df$MaxTemp))
sum(is.na(rain_df$Rainfall))
sum(is.na(rain_df$Evaporation))
sum(is.na(rain_df$Sunshine))
sum(is.na(rain_df$WindGustDir))
sum(is.na(rain_df$WindGustSpeed))
sum(is.na(rain_df$WindDir9am))
sum(is.na(rain_df$WindDir3pm))
sum(is.na(rain_df$WindSpeed9am))
sum(is.na(rain_df$WindSpeed3pm))
sum(is.na(rain_df$Humidity9am))
sum(is.na(rain_df$Humidity3pm))
sum(is.na(rain_df$Pressure9am))
sum(is.na(rain_df$Pressure3pm))
sum(is.na(rain_df$Cloud9am))
sum(is.na(rain_df$Cloud3pm))
sum(is.na(rain_df$Temp9am))
sum(is.na(rain_df$Temp3pm))
sum(is.na(rain_df$RainToday))
sum(is.na(rain_df$RainTomorrow))
```

```{r}
# Chgeck the amount of data after dropping sunshine variable na

dim(rain_df)
```

## 2.5. Handling the outliers

### Rainfall

```{r}
# Check for outliers using Box Plot

boxplot(Rainfall ~ RainTomorrow, data = rain_df) 
```

```{r}
# Check on histogram

hist(rain_df$Rainfall)
```

```{r}
# Use IQR to define range of valid data 

Q1 <- quantile(rain_df$Rainfall, 0.25)
Q3 <- quantile(rain_df$Rainfall, 0.75)

IQR <- Q3 - Q1

Low <- Q1 - IQR * 1.5
High <- Q3 + IQR * 1.5

print(Low)
print(High)
```

```{r}
# Check the amount of outliers 

count1 <- sum(rain_df$Rainfall > High)
count2 <- sum(rain_df$Rainfall < Low)
count3 <- count1 + count2

print(count1)
print(count2)
```

```{r}
# We decided to censor our data to value of 75, because our variable is not normally distributed (high density of data near 0)

rain_df$Rainfall[rain_df$Rainfall > 75] <- 75
```

```{r}
# Check changes on histogram

hist(rain_df$Rainfall)
```

```{r}
# Check Box Plot again

boxplot(Rainfall ~ RainTomorrow, data = rain_df)
```

### Evaporation

```{r}
# Check for outliers using Box Plot

boxplot(Evaporation ~ RainTomorrow, data = rain_df) 
```

```{r}
# Check on histogram

hist(rain_df$Evaporation)
```

```{r}
# Use IQR to define range of valid data 

Q1 <- quantile(rain_df$Evaporation, 0.25)
Q3 <- quantile(rain_df$Evaporation, 0.75)

IQR <- Q3 - Q1

Low <- Q1 - IQR * 1.5
High <- Q3 + IQR * 1.5

print(Low)
print(High)
```

```{r}
# Check the amount of outliers 

count1 <- sum(rain_df$Evaporation > High)
count2 <- sum(rain_df$Evaporation < Low)
count3 <- count1 + count2

print(count1)
print(count2)
```

```{r}
# Change outliers respectivley to upper boundry since (relativley) there is not too many outliers

rain_df$Evaporation[rain_df$Evaporation > High] <- High
```

```{r}
# Check changes on histogram

hist(rain_df$Evaporation)
```

```{r}
# Check Box Plot again

boxplot(Evaporation ~ RainTomorrow, data = rain_df)
```

### WindGustSpeed

```{r}
# Check for outliers using Box Plot

boxplot(WindGustSpeed ~ RainTomorrow, data = rain_df) 
```

```{r}
# Check on histogram

hist(rain_df$WindGustSpeed)
```

```{r}
# Use IQR to define range of valid data 

Q1 <- quantile(rain_df$WindGustSpeed, 0.25)
Q3 <- quantile(rain_df$WindGustSpeed, 0.75)

IQR <- Q3 - Q1

Low <- Q1 - IQR * 1.5
High <- Q3 + IQR * 1.5

print(Low)
print(High)
```

```{r}
# Check the amount of outliers 

count1 <- sum(rain_df$WindGustSpeed > High)
count2 <- sum(rain_df$WindGustSpeed < Low)
count3 <- count1 + count2

print(count1)
print(count2)
```

```{r}
# Change outliers respectivley to upper boundry since (relativley) there is not too many outliers

rain_df$WindGustSpeed[rain_df$WindGustSpeed > High] <- High
```

```{r}
# Check changes on histogram

hist(rain_df$WindGustSpeed)
```

```{r}
# Check Box Plot again

boxplot(WindGustSpeed ~ RainTomorrow, data = rain_df)
```

### WindSpeed9am


```{r}
# Check for outliers using Box Plot

boxplot(WindSpeed9am ~ RainTomorrow, data = rain_df) 
```

```{r}
# Check on histogram

hist(rain_df$WindSpeed9am)
```

```{r}
# Use IQR to define range of valid data 

Q1 <- quantile(rain_df$WindSpeed9am, 0.25)
Q3 <- quantile(rain_df$WindSpeed9am, 0.75)

IQR <- Q3 - Q1

Low <- Q1 - IQR * 1.5
High <- Q3 + IQR * 1.5

print(Low)
print(High)
```

```{r}
# Check the amount of outliers 

count1 <- sum(rain_df$WindSpeed9am > High)
count2 <- sum(rain_df$WindSpeed9am < Low)
count3 <- count1 + count2

print(count1)
print(count2)
```

```{r}
# Change outliers respectivley to upper boundry since (relativley) there is not too many outliers

rain_df$WindSpeed9am[rain_df$WindSpeed9am > High] <- High
```

```{r}
# Check changes on histogram

hist(rain_df$WindSpeed9am)
```

```{r}
# Check Box Plot again

boxplot(WindSpeed9am ~ RainTomorrow, data = rain_df)
```

### WindSpeed3pm

```{r}
# Check for outliers using Box Plot

boxplot(WindSpeed3pm ~ RainTomorrow, data = rain_df) 
```

```{r}
# Check on histogram

hist(rain_df$WindSpeed3pm)
```

```{r}
# Use IQR to define range of valid data 

Q1 <- quantile(rain_df$WindSpeed3pm, 0.25)
Q3 <- quantile(rain_df$WindSpeed3pm, 0.75)

IQR <- Q3 - Q1

Low <- Q1 - IQR * 1.5
High <- Q3 + IQR * 1.5

print(Low)
print(High)
```

```{r}
# Check the amount of outliers 

count1 <- sum(rain_df$WindSpeed3pm > High)
count2 <- sum(rain_df$WindSpeed3pm < Low)
count3 <- count1 + count2

print(count1)
print(count2)
```

```{r}
# Change outliers respectivley to upper boundry since (relativley) there is not too many outliers

rain_df$WindSpeed3pm[rain_df$WindSpeed3pm > High] <- High
```

```{r}
# Check changes on histogram

hist(rain_df$WindSpeed3pm)
```

```{r}
# Check Box Plot again

boxplot(WindSpeed3pm ~ RainTomorrow, data = rain_df)
```

There was a lot of outliers in variables **Rainfall, Evaporation, WindGustSpeed, WindSpeed9am, WindSpeed3pm** after further investigation we decided to change the outstanding values, using IQR in order to determine the maximal and minimal value we changed the data respectivley.

## 2.6. Data Simplification

#```{r}
# Since the wind direction were wide spread and very specific we decided to simplify the data by puting them into categories based on dominant direction, filled the missing data for wind direction using mode.

rain_df$WindDir9am <- replace("NNW", rain_df$WindDir9am, "N")


rain_df$WindDir9am <- ifelse(is.na(rain_df$WindDir9am), mode(rain_df$WindDir9am)[1], rain_df$WindDir9am)
rain_df$WindDir3pm <- ifelse(is.na(rain_df$WindDir3pm), mode(rain_df$WindDir3pm)[1], rain_df$WindDir3pm)
rain_df$WindGustDir <- ifelse(is.na(rain_df$WindGustDir), mode(rain_df$WindGustDir)[1], rain_df$WindGustDir)

tail(rain_df)
#```

## 2.7. Handling Date

```{r}
n_distinct(rain_df$Date)
```

```{r}
# Change into datetime format in order to extarct month.

rain_df$Date <- as.POSIXct(rain_df$Date)
rain_df$Date <- month(rain_df$Date)
names(rain_df)[names(rain_df) == "Date"] <- "Month"

tail(rain_df)
```

# 3. Selecting Features

## 3.1. Creating new numerical variables

NADIMS + MICHALS PART

# 4. Balancing and Standarizing

NADIMS PART

# 5. Split train and test

#```{r}
set.seed(111)

split1<- sample(nrow(rain_df),floor(nrow(rain_df)*0.7),replace=FALSE)

train <- rain_df[split1,]

test <- rain_df[-split1,]
#<<<<<<< HEAD
#```
=======
head(train)
```
>>>>>>> 1b3e2b3bbf38c00b31d322a3bd672e29a047b39a

# 6. Logistic Regression
#```{r}
logistic_reg <- glm(RainTomorrow~.,family = binomial(link = "logit"),data=train)
```
We won't be doing hyperparameter tuning for logistic regression, as there is not much it could be done.

# 7. SVC
#```{r}
SVM_model <- svm(RainTomorrow~., data=train)

SVM_tuned <- tune.svm(RainTomorrow~., data=train,
                      degree = c(3,4),
                      gamma = c(0.001 ,.005 ,0.01 ,0.1),
                      cost = seq(0.1,1, by = 0.1),
                      epsilon = c(0.1,0.05)
                      )
print(SVM_tuned$best.performance)
SVM_tuned$best.model
```
# 8. Random Forest

NADIMS PART

# 9. xgboost

NADIMS PART

# 10. Conclusion

MICHALS PART
